Beyond the Arc: The Impact of  3 point Shooting in the Modern NBA

### 2024-04-18 paper/paper.qmd
Update paper.qmd

Understanding this link has ramifications for numerous parties, including coaches, players, analysts, and team administration. This study adds to the ongoing debate in the NBA about strategic decision-making and player development by explaining the significance of three-point shooting in deciding game outcomes (@cite-debate). Essentially, the question “Does three point shooting actually lead to team success?”, is being asked.

@@

## Data Collection

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-1
#| tbl-cap: "Overview of the cleaned dataset"
#| warning: false

@@

In this study, our focus is solely on win counts and three-point statistics. With this objective in mind, our key variables of interest include Games Played (**GP**), Wins (**W**), Losses (**L**), Three-Pointers Made per game (**3PM**), Three-Pointers Attempted per game (3PA), and Three-Point Shooting Percentage (**3P%**). Our modeling approach will specifically incorporate Wins (**W**) and **3PM** to explore their relationship.

@@

By prioritizing **3PM** alongside Wins, our model aims to uncover insights into the impact of three-point scoring on overall team success while mitigating potential biases associated with alternative metrics. This strategic selection ensures a throrough exploration of the relationship between three-point performance and winning outcomes in NBA games.

@@

```{r}
#| echo: false
#| label: fig-2
#| fig-cap: "Scatterplot depicting the **Wins** vs the **3 Points Made** per game"
#| warning: false
ggplot(cleaned_data, aes(x = X3PM, y = W)) +
  geom_point(color = "orange") +
  labs(x = "3PM", y = "Wins", title = "Wins vs 3PM") +
  theme_minimal() +  # Apply a minimal theme
  theme(
    axis.line = element_line(color = "black"),  # Add axis lines
    panel.grid.major = element_blank(),         # Remove major gridlines
    panel.grid.minor = element_blank()          # Remove minor gridlines
  )
```

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-3
#| tbl-cap: "Table showing the win stats and 3 point stats grouped by the Team"
#| warning: false

@@

# Display the sorted table
kable(sorted_table)
```
```{r}
#| echo: false
#| eval: true
#| label: fig-3
#| fig-cap: "Graph showing the average 3 point shots made per game across all the seasons"
#| warning: false

team_colors <- c(
  "Atlanta Hawks" = "#E03A3E",
  "Boston Celtics" = "#007A33",
  "Brooklyn Nets" = "#000000",
  "Charlotte Hornets" = "#00788C",
  "Chicago Bulls" = "#CE1141",
  "Cleveland Cavaliers" = "#860038",
  "Dallas Mavericks" = "#00538C",
  "Denver Nuggets" = "#0E2240",
  "Detroit Pistons" = "#C8102E",
  "Golden State Warriors" = "#1D428A",
  "Houston Rockets" = "#CE1141",
  "Indiana Pacers" = "#002D62",
  "LA Clippers" = "#C8102E",
  "Los Angeles Lakers" = "#552583",
  "Memphis Grizzlies" = "#5D76A9",
  "Miami Heat" = "#98002E",
  "Milwaukee Bucks" = "#00471B",
  "Minnesota Timberwolves" = "#0C2340",
  "New Orleans Pelicans" = "#0C2340",
  "New York Knicks" = "#006BB6",
  "Oklahoma City Thunder" = "#007AC1",
  "Orlando Magic" = "#0077C0",
  "Philadelphia 76ers" = "#006BB6",
  "Phoenix Suns" = "#1D1160",
  "Portland Trail Blazers" = "#E03A3E",
  "Sacramento Kings" = "#5A2D81",
  "San Antonio Spurs" = "#C4CED4",
  "Toronto Raptors" = "#CE1141",
  "Utah Jazz" = "#002B5C",
  "Washington Wizards" = "#002B5C"
)
ggplot(grouped_data, aes(x = Average_3PM, y = reorder(TEAM, Average_3PM), fill = TEAM)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Average_3PM, 1)), position = position_nudge(x = 0.5), vjust = 0.5, size = 3, color = "black") +
  labs(x = "Average 3-Pointers Made", y = "Team") +
  scale_fill_manual(values = team_colors) +
  theme(legend.position = "none", axis.text.y = element_text(size = 8))
```
@fig-3 shows the average 3 point shots made per game across the 8 seasons that are involved in our study.
```{r}
#| echo: false
#| label: fig-1
#| fig-cap: "Graph depicting the **Average team wins** and **predicted team wins** vs **Average three point shots made** per game"
#| warning: false
# Calculate predicted wins
predicted_wins <- fitted(nba_model2)

@@

Although, this paper discusses the upside of the three point line, I feel it is important to also include it's drawbacks. An interesting example of this in the modern NBA is the 2017-18 Houston Rockets. The Houston Rockets' general manager was a huge fan of the three point shot and in the 2017-18 season built an entire roster around 3 point shooting (@cite-Rockets1). This team was a huge success in the regular season and they shot more 3 pointers than any other team in league history. What he failed to take into account was, that basketball isn't math. While the Rockets were the most successful regular season team with 65 wins (@cite-Rockets2), they matched up against the Golden State Warriors in the western conference finals of the playoffs. The series went to seven games, and in game 7, the Rockets essentially shot themselves out of the game. They went 7 for 49 from the arc as a team, which was a record for the worst 3 point shooting in NBA history. This team lived by the three and eventually would die by the three.

>>>>

Understanding this link has ramifications for numerous parties, including coaches, players, analysts, and team administration. This study adds to the ongoing debate in the NBA about strategic decision-making and player development by explaining the significance of three-point shooting in deciding game outcomes (@cite-debate). Essentially, the estmand of the paper is to see if an increase in a team's three point shooting ability contributes to an increase in the team's success, measured by wins.

@@

## Data Sources

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-1
#| tbl-cap: "Summary table of the cleaned dataset"
#| warning: false

@@

In this study, our focus is solely on win counts and three-point statistics. With this objective in mind, our key variables of interest include Games Played (**GP**), Wins (**W**), Losses (**L**), Three-Pointers Made per game (**3PM**), Three-Pointers Attempted per game (**3PA**), and Three-Point Shooting Percentage (**3P%**). Our modeling approach will specifically incorporate Wins (**W**) and **3PM** to explore their relationship.

@@

By prioritizing **3PM** alongside Wins, our model aims to uncover insights into the impact of three-point scoring on overall team success while mitigating potential biases associated with alternative metrics. This strategic selection ensures a detailed exploration of the relationship between three-point performance and winning outcomes in NBA games.

@@

```{r}
#| echo: false
#| label: fig-2
#| fig-cap: "Scatterplot depicting the **Wins** vs the **3 Points Made** per game by all teams across the seasons 2014-15 to 2022-23"
#| warning: false
ggplot(cleaned_data, aes(x = X3PM, y = W)) +
  geom_point(color = "orange") +
  labs(x = "3PM", y = "Wins", title = "Wins vs 3PM") +
  theme_minimal() +  # Apply a minimal theme
  theme(
    axis.line = element_line(color = "black"),  # Add axis lines
    panel.grid.major = element_blank(),         # Remove major gridlines
    panel.grid.minor = element_blank()          # Remove minor gridlines
  )
```

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-3
#| tbl-cap: "Summary of Team statistics across seasons 2014-15 through 2022-23"
#| warning: false

@@

# Display the sorted table
kable(sorted_table)
```
```{r}
#| echo: false
#| eval: true
#| label: fig-3
#| fig-cap: "Visualization of Average 3-Pointers Made per Game Across NBA Seasons"
#| warning: false

team_colors <- c(
  "Atlanta Hawks" = "#E03A3E",
  "Boston Celtics" = "#007A33",
  "Brooklyn Nets" = "#000000",
  "Charlotte Hornets" = "#00788C",
  "Chicago Bulls" = "#CE1141",
  "Cleveland Cavaliers" = "#860038",
  "Dallas Mavericks" = "#00538C",
  "Denver Nuggets" = "#0E2240",
  "Detroit Pistons" = "#C8102E",
  "Golden State Warriors" = "#1D428A",
  "Houston Rockets" = "#CE1141",
  "Indiana Pacers" = "#002D62",
  "LA Clippers" = "#C8102E",
  "Los Angeles Lakers" = "#552583",
  "Memphis Grizzlies" = "#5D76A9",
  "Miami Heat" = "#98002E",
  "Milwaukee Bucks" = "#00471B",
  "Minnesota Timberwolves" = "#0C2340",
  "New Orleans Pelicans" = "#0C2340",
  "New York Knicks" = "#006BB6",
  "Oklahoma City Thunder" = "#007AC1",
  "Orlando Magic" = "#0077C0",
  "Philadelphia 76ers" = "#006BB6",
  "Phoenix Suns" = "#1D1160",
  "Portland Trail Blazers" = "#E03A3E",
  "Sacramento Kings" = "#5A2D81",
  "San Antonio Spurs" = "#C4CED4",
  "Toronto Raptors" = "#CE1141",
  "Utah Jazz" = "#002B5C",
  "Washington Wizards" = "#002B5C"
)
ggplot(grouped_data, aes(x = Average_3PM, y = reorder(TEAM, Average_3PM), fill = TEAM)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Average_3PM, 1)), position = position_nudge(x = 0.5), vjust = 0.5, size = 3, color = "black") +
  labs(x = "Average 3-Pointers Made", y = "Team") +
  scale_fill_manual(values = team_colors) +
  theme(legend.position = "none", axis.text.y = element_text(size = 8))
```
@fig-3 shows the average 3 point shots made per game across the 8 seasons that are involved in our study.
```{r}
#| echo: false
#| label: fig-1
#| fig-cap: "Relationship Between **Average Team Wins**, **Predicted Wins**, and **Average Three-Point Shots Made per Game**"
#| warning: false
# Calculate predicted wins
predicted_wins <- fitted(nba_model2)

@@

Although this paper discusses the upside of the three point line, I feel it is important to also include its drawbacks. An interesting example of this in the modern NBA is the 2017-18 Houston Rockets. The Houston Rockets' general manager was a huge fan of the three point shot and in the 2017-18 season built an entire roster around 3 point shooting (@cite-Rockets1). This team was a huge success in the regular season and they shot more 3 pointers than any other team in league history. What he failed to take into account was, that basketball isn't math. While the Rockets were the most successful regular season team with 65 wins (@cite-Rockets2), they matched up against the Golden State Warriors in the western conference finals of the playoffs. The series went to seven games, and in game 7, the Rockets essentially shot themselves out of the game. They went 7 for 49 from the arc as a team, which was a record for the worst 3 point shooting in NBA history. This team lived by the three and eventually would die by the three.
